### 2019-09-06 Manuscripts/cogsci-journal/cogsci-journal.tex
Cog Sci manuscript finalized, and ready for submission.

\vfill
Joseph L. Austerweil\textsuperscript{1}
Nolan Conaway\textsuperscript{1}, 
Shi Xian Liew\textsuperscript{1},
and Kenneth J. Kurtz\textsuperscript{2}
\\[\baselineskip]
\textsuperscript{1}Department of Psychology, University of Wisconsin-Madison, Madison, WI, USA
\textsuperscript{2}Department of Psychology, Binghamton University, Binghamton, NY, USA
\\[1in]

@@

Correspondence concerning this article should be addressed to: 
Joseph L. Austerweil, 1202 West Johnson Street, Madison, WI 53706.
E-mail: austerweil@wisc.edu

@@

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Intuitively, the influence of contrast may be expected in a number of
real-life scenarios involving the generation of new categories. For instance, a
college student who has been eating instant noodles for a week may be more
inclined to seek out something healthier and less carbohydrate-heavy (e.g., a
Caesar salad) compared to something similar (a different flavour of instant
noodles). An actor who is dissatisfied with being typecast as a villain may seek
out roles as heroes as opposed to playing other antagonistic characters. An
especially compelling example can be seen in \cite{askin2017} who found that
new songs that were more different to their peers were generally more popular
than new songs that were similar to others. It appears that once there is the
motivation to generate something new, there is an associated motivation to
generate something \emph{different}.

@@

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic \citep{kahneman1972subjective}
states that people make judgments regarding an outcome based on how
representative it is of the evidence given in the current context. The heuristic
is a powerful theoretical construct that has been used to capture a range of
complex patterns of human judgments \citep{kahneman73,tversky74,tversky83},
especially those that deviate from normative theory (as given by a
straightforward application of Bayes' rule). For example, the coinflip sequence
TTHTH is perceived to be more random (due to it being representative of a
coinflip sequence) than TTTTT (despite both having the same probability of
having been generated from a fair coin). Although there is a healthy debate as
to whether perceived randomness {\em really} deviates from normative theory
\citep{griffiths18,hahn09}, the representativeness heuristic remains one of the
main explanations of human judgments \citep{reimers18}. To formalize the
hypothesis that people expect a new category to be representative of the
opposite of the current category, or contrast as {\em representativeness}, we
use the Bayesian formulation of representativeness \citep{tenenbaum01}, which
has been used successfully to capture human performance in color categorization
\citep{abbott16}, image category learning \citep{abbott11}, and language grammar
learning \citep{rafferty10}. We do so by extending the state-of-the-art category
generation model \citep{jern2013probabilistic} and find that human category
generation is best captured as generating representative, rather than probable,
samples.

@@

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over feature space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been throughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns how contrast may be captured through the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}
% 
%where $\theta$ ($\theta \geq 0$) is a free parameter controlling response
%determinism.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer}
%: Both models are
%exemplar-based, and both models generate new items according to their similarity
%to known members of the target class. However, PACKER diverges from the
%copy-and-tweak model by including a contrast mechanism, enabling generation
%according to dissimilarity to members of opposing categories. As a consequence,
%copy-and-tweak can be realized as a parameterization of the PACKER model that is
%insensitive to category contrast. Specifically, when $\theta_c = 0$ and
%$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
%$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
%items, and is mathematically equivalent to a copy-and-tweak approach.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories.
The general idea of representativeness is not new in categorization. For
instance, the literature on the \emph{graded structure} of categories explores
how some exemplars are better representations of their categories
\citep{palmeri2001central,barsalou1985ideals}. In this article we adopt the
specific formalization of representativeness given by \cite{tenenbaum01}, where
the representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is
the relative amount of evidence that is provided by $y$ for a given hypothesis
$h$ in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

@@

183 participants were recruited from Amazon Mechanical Turk. Each participant
was randomly assigned to one condition: 64 participants were assigned to the
Cluster condition, 61 were assigned to the Row condition, and 58 were assigned
to the XOR condition (sample sizes differ due to random assignment). Stimuli
were squares varying in color (grayscale 9.8\%--90.2\%) and side length
(3.0--5.8cm), see Figure \ref{fig:e1-conditions}. The assignment of perceptual
features (color, size) to axes of the conceptual space (x, y)  and the
direction of variation along each axis (e.g., $dark \rightarrow light$ or $light
\rightarrow dark$) were counterbalanced across participants.

@@

A secondary goal of this experiment was to examine whether we replicate the
classic result that generated categories often possess the same distributional
properties as previously-known categories. Given the increased importance of replication within psychology \citep{zwaan2018making}, it is important as it serves as a conceptual replication of \citet{jern2013probabilistic}. For each generated category, we
computed the category range along each axis (X, Y), as well as the correlation
between features. These data, shown in Figure \ref{fig:e1-statsboxes}, reveal
broad individual differences: Within each condition, participants generated
categories spanning the entire X- and Y- axis as well as categories that spanned
very little along each. Likewise, in each condition participants generated
categories possessing strongly positive, neutral, and strongly negative
correlations between the dimensions. Comparing the distributional statistics
between conditions yields a broad yet, as we will see, misleading replication of
the classic effect.

@@

However, our ability to detect differences in Y-Axis range using a standard
$t$-test between the conditions is, in this case, diminished due to the
non-normality of the data (Shapiro-Wilk normality test $W=0.77, p < .001$ for
the Middle condition and $W=0.85, p < .001$ for the Bottom condition). Figure \ref{fig:e2-yranges} depicts the Y-Axis range and Y-Axis position of exemplars generated by each participant. The categories are sorted by overall range, and then colored by training condition. These data reveal that there were nearly as many
participants who generated categories spanning the entire Y-axis as those who
generated categories spanning almost none of the Y-axis. Indeed, the median produced Y-Axis range is much smaller in the Bottom than Middle condition, whereas they are essentially identical in X-Axis range. The non-normality of the Y-axis range distributions thus requires that we use a different analytic approach to
addressing the experiment's main question.

\begin{figure}
    \begin{center} \inputpgf{figs/}{e2-yranges.pgf}
    \caption{Y-Axis range and position of the participant-generated categories
from Experiment 2. Each line corresponds to a participant's category, with
notches corresponding to the Y-axis position of exemplars within the category
(notches may overlap). Participants are sorted by overall range, and then by
condition. }
    \label{fig:e2-yranges}
    \end{center}
\end{figure}

@@

To obtain a global measure of the quality of each model's account, we fitted the
parameters of each model to our entire dataset (Experiments 1 and 2 combined),
using a hill-climbing algorithm which maximized the log-likelihood of the
model's predictions of the observed responses (1220 responses from 305 total
participants). We fitted three parameters in the PACKER model ($c$, $\theta_t$,
and $\theta_c$; see Section \ref{section:PACKER-definition}), as well as four in
the representativeness model and the hierarchical Bayesian model ($\kappa$,
$\lambda$, $\nu$, and $\theta$; see Section
\ref{section:representativeness-definition} and Appendix
\ref{ap:hsampling-definition} respectively). We fitted only two parameters for
the copy-and-tweak model ($c$, and $\theta_t$), as $\theta_c$ is held constant
($\theta_c = 0$). Attention ($w$, see Equation \ref{eq:similarity}) in PACKER
and copy-and-tweak was set uniformly. Parameters were not allowed to vary
between participants or conditions -- the goal was to obtain the best-fitting
values to our entire dataset.

@@

While PACKER's advantage may tentatively be attributed to the model's
sensitivity to category contrast (this will be explored in detail below), the
advantage shown by copy-and-tweak over the hierarchical Bayesian model may be
attributed to its exemplar-based representation of category B, as opposed to forcing a
prototype-based representation as assumed by the hierarchical Bayesian model. As
observed in Figures \ref{fig:e1-samples} and \ref{fig:e2-samples}, the generated
categories we observed were often widely distributed, with no items near the
category prototype. This aspect of the data is inconsistent with the
multivariate normal distributions (similar to prototypes) used to represent
categories in the \cite{jern2013probabilistic} model, but can be easily
accounted for using an exemplar-based approach. Interestingly, representativeness using a prototype approach fits better than an exemplar-based approach. Understanding these differences more thoroughly is left for future research.

@@

In this section, we focus on a different category type -- the Corner category -- that contrasts the two models of
contrast. Here, each Alpha exemplar is located at the corners of the feature space (see Figure \ref{fig:e3-squares}a).

We first identify the predictions of each model in this condition. Second, we test these contrasting predictions in a
behavioral experiment. We then conduct a model-based analysis of the two contrast models.

@@

Thus, in this experiment, we also examine how making a boundary toroidal (e.g., line rotation) affects the prediction of
each model. The representativeness model prefers to generate categories on the feature that still has a boundary.
PACKER, in contrast, should continue to generate categories that are most distant from the learn category regardless of
the presence or absence of feature boundaries -- in this condition it would be the center of the feature space.
Experiment 3 manipulates whether one feature is bounded or toroidal (always displayed as the Y-axis). One condition uses
the same stimuli as before, squares varying in lightness and size, both of which are bounded features. The other
condition use a new set of stimuli: circles that vary in size (a bounded feature) and orientation (an unbounded feature;
Figure \ref{fig:e3-circles}a). Before describing the details of the experiment, we explain in detail each model's
predictions given the Corner Alphas for a feature space where one feature is bounded and the other is toroidal.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-circles.pdf}
      \caption{Circle stimulus domain (a) and predictions from contrast models
        PACKER (b) and the representativeness model (c). The unbounded feature
        (orientation) is represented on the y-axis, while the bounded feature
        (size) is represented on a x-axis. As with the square stimuli, feature
        and direction assignment were counterbalanced across participants (in
        the displayed example, circles are rotated clockwise up the y-axis). }
      \label{fig:e3-circles}
    \end{center}
\end{figure}

@@

For the representativeness model, adding a toroidal feature changes the probability distribution for the underlying
Bayesian model. For the toroidal feature, space wraps back to $0$ radians once it reaches $2\pi$ (we have conducted a
change of variables where $y \rightarrow y$ mod $2\pi$, where mod is the modulus operator). As we cross $2\pi$ and reach
$0$ again, the remaining probability density also gets wrapped around, meaning that the density at $2\pi$ is added to
$0$ (and $2\pi + \epsilon$ to $\epsilon$, and so on). This is called the {\em wrapped} Gaussian distribution
\citep{mardia1972}. Note that the more familiar distribution for circular features, the Von Mises distribution, is an
approximation to the wrapped Gaussian distribution. See Appendix B for implementational details of the
representativeness model in a bounded and toroidal feature.

@@

% It may be argued that our expectation that the representativeness model should
% \emph{always} tend to produce first-Betas at the edges may be too strong.
% Specifically, this prediction is based on the model that is optimized to data
% from Experiments 1 and 2, and it is conceivable that the model may make
% drastically different predictions (such as some favoring the initial generation
% of centrally-located Betas more than Betas at the edges) in the current
% experiment when it is actually fit to empirical data from this experiment.
% However, we contend that this prediction of the representativeness model is a
% fundamental property of the model due to its assumption of unimodal category
% distributions. At the very least, the maximum probability of generating a
% centrally located first Beta should almost never be higher than the maximum
% probability of generating a first Beta at an edge.

% It is also worth noting that since the representativeness model does not explicitly
% recognize similarities as distances within the feature space (unlike PACKER),
% there is no compression in density regions along the unbounded feature axis.
% This provides us with an additional qualitative marker distinguishing the
% representativeness model from PACKER -- if the range of the entire Beta category
% along the unbounded feature is observed to be shorter than that of the bounded
% feature, this would be stronger evidence for PACKER rather than the
% representativeness model.

% In the following
% section, we explore -- using both qualitative comparisons and quantitative fits
% -- whether PACKER or the representativeness model is a better account of
% behavioral data in this Corner experiment.

% For the representativeness model, the single unbounded feature essentially
% prevents the formation of edge effects along that feature. Consequently, this results in higher probabilities of the generating the first Beta exemplars along the edges of the bounded feature (Figure \ref{fig:e3-circles}c). As this
% model does not explicitly recognize similarities as distances along the feature
% space (unlike PACKER), it does not experience the compression in density regions
% along the unbounded feature axis. Intuitively, this suggests that the
% representativeness model (which prefers generating exemplars at the boundaries) will prefer to generate stimuli on the boundaries of the bounded feature with some amount of variance over unbounded feature.

% We have identified the conditions under which the contrast models may make
% differing predictions, however this distinction is not very useful without any
% empirical support. Data from the next experiment will most certainly support one
% model or the other

%[Maybe work on transition if there's time]

@@

We recruited 89 participants from Amazon Mechanical Turk who were randomly assigned to the square stimuli ($N=46$), or
to the circle stimuli ($N=43$). The square stimuli were exactly the same as in Experiments 1 and 2, but with Alpha
exemplars in the corners of the feature space. The circle stimuli were made up of unfilled circles with a radial line
extending from its center to its edge. These stimuli varied along two features -- their diameter ranging from 3.0--5.8
cm, and their orientation ranging from $2.587$--$2\pi + 2.587$ radians, with $0$ radians indicating a completely
horizontal line. The line widths of the circles and radial line were set to 0.05 cm. Similar to Experiments 1 and 2,
participants first observed a training phase where four Alpha exemplars were presented across three blocks followed by
the generation phase, and the features values are divided into 9 discrete steps. The training condition was the Corner
category for all participants.

@@

The distributions of generated first-Betas (Figure \ref{fig:e3-behave}) indicate further qualitative support for PACKER.
In particular, the probability of generating the first Beta exemplar for circles along the bounded dimension is
significantly negatively correlated with distance from the center ($r(3) = -.95, p = .01$). Although the correlations
for squares ($r(3)= -.50, p = .40$ and $r(3)= -.31 p = .61$ for x and y dimensions) were not significant, they were in
the directions predicted by PACKER. Mean probabilities of of generating first-Betas are presented in Figure
\ref{fig:e3-distcent}. This analysis was excluded for the toroidal dimension of the circle stimuli because the
boundaries (and consequently a center) are not defined for this feature.

% An inspection of the distributions of generated first-Betas (Figure
% \ref{fig:e3-behave}) does not immediately indicate support for either model. The
% square condition reveals first Betas generated at the corners as well as the
% center of the feature space, while the circle condition shows greater
% frequencies at the center. However, we also found that 65\% and 87\% of
% participants generated their first squares and first circles, respectively, away
% from the edge of the feature space.\footnote{Strictly speaking, the circle
%   stimuli do not have edges along its orientation feature, However, for ease of
%   comparison we will continue to refer to these comparable regions between
%   stimuli as edges.} Further, the maximum proportion of first-Betas generated
% off the edges (.13) is higher than that of the maximum proportion in an edge
% (.04) for the circle stimuli, although the maximum proportions are the same for
% squares (.06 for both edge and off-edge). This preference for generating
% first-Betas away from the edge is in line with our hypothesis
% supporting PACKER.

@@

\begin{figure}
    \begin{center}
      \includegraphics[width=.6\textwidth]{figs/e3-distcent.pdf}
      \caption{Empirical mean probabilities of generating first Beta exemplars at each point of distance from the center
        of the feature space. Plot for the toroidal feature of circle stimuli is not presented because a center is not
        defined for this feature.}
    \label{fig:e3-distcent}
    \end{center}
\end{figure}

>>>>

\vfill
Joseph L. Austerweil\textsuperscript{1+}
Shi Xian Liew\textsuperscript{1+},
Nolan Conaway\textsuperscript{2}, 
and Kenneth J. Kurtz\textsuperscript{3}
\\[\baselineskip]
\textsuperscript{1}Department of Psychology, University of Wisconsin-Madison, Madison, WI, USA\\
\textsuperscript{2}Shutterstock, New York, NY, USA\\
\textsuperscript{3}Department of Psychology, Binghamton University, Binghamton, NY, USA
\\[1in]

@@

Correspondence concerning this article should be addressed to: 
Shi Xian Liew, 1202 West Johnson Street, Madison, WI 53706.
E-mail: shixianliew@gmail.com

\textsuperscript{+}These authors share equal contribution to this work.

@@

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Intuitively, the influence of contrast may be expected in a number of
real-life scenarios involving the generation of new categories. For instance, a
college student who has been eating instant noodles for a week may be more
inclined to seek out something healthier and less carbohydrate-heavy (e.g., a
Caesar salad) compared to something similar (a different flavor of instant
noodles). An actor who is dissatisfied with being typecast as a villain may seek
out roles as heroes as opposed to playing other antagonistic characters. An
especially compelling example can be seen in \cite{askin2017} who found that new
songs that were more different to their peers were generally more popular than
new songs that were similar to others. It appears that once there is the
motivation to generate something new, there is an associated motivation to
generate and observe something \emph{different}.

@@

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic
\citep{kahneman1972subjective} states that people make judgments regarding an
outcome based on how representative it is of the evidence given in the current
context. The heuristic is a powerful theoretical construct that has been used to
capture a range of complex patterns of human judgments
\citep{kahneman73,tversky74,tversky83}, especially those that deviate from
normative theory (as given by a straightforward application of Bayes' rule). For
example, the coinflip sequence TTHTH is perceived to be more random (due to it
being representative of a coinflip sequence) than TTTTT (despite both having the
same probability of having been generated from a fair coin). Although there is a
healthy debate as to whether perceived randomness {\em really} deviates from
normative theory \citep{griffiths18,hahn09}, the representativeness heuristic
remains one of the main explanations of human judgments \citep{reimers18}. To
formalize the hypothesis that people expect a new category to be representative
of the opposite of the current category, or contrast as {\em
  representativeness}, we use the Bayesian formulation of representativeness
\citep{tenenbaum01}, which has been used successfully to capture human
performance in color categorization \citep{abbott16}, image category learning
\citep{abbott11}, and language grammar learning \citep{rafferty10}. We do so by
extending the state-of-the-art category generation model
\citep{jern2013probabilistic} and find that human category generation can
also be captured as generating representative, rather than probable, samples.

@@

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over feature space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generate categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting exemplars from a new category.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been thoroughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns how contrast may be captured through the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

@@

The probability that a candidate $y$ will be generated is evaluated using an
Exponentiated \citet{luce1977choice} choice rule. Candidates with greater values
of $a(y,x)$ are more likely to be generated than candidates with smaller values:
% 
\begin{equation} p(y \mid x) = \dfrac { \exp \left \{ a \left( y, x
\right) \right \} } { \sum_i{ \exp \left \{ a \left( y_i, x \right)
\right\} } }
\label{eq:packer-choice}
\end{equation}

Note, because $\theta_t$ and $\theta_c$ are unconstrained, their size plays the
role of the response determinism parameter for the exponentiated Luce choice
rule.

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer}

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories.
The general idea of representativeness is not new in categorization. For
instance, the literature on the \emph{graded structure} of categories explores
how some exemplars are better exemplars of their categories
\citep{palmeri2001central,barsalou1985ideals}. In this article we adopt the
specific formalization of representativeness given by \cite{tenenbaum01}, where
the representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is
the relative amount of evidence that is provided by $y$ for a given hypothesis
$h$ in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$ (i.e., we effectively set
$p(h)$ to 0 and then renormalize the priors). Note that when there are only two
hypotheses (e.g. two categories) involved in the domain, this prior takes the
value of $1$ since there is ever only one alternative hypothesis.

@@

Here, we implement the representativeness mechanism at the level of
distributions over category exemplars. However, with a hierarchical model, it is
also possible to apply the representativeness mechanism at other levels, such as
the level of the distribution of category statistics (i.e., their means and
covariances). For simplicity, we focus only on one type of representativeness
model (i.e., where the mechanism is implemented at the level of distribution
over category exemplars).

@@

183 participants were recruited from Amazon Mechanical Turk. Each participant
was randomly assigned to one condition: 64 participants were assigned to the
Cluster condition, 61 were assigned to the Row condition, and 58 were assigned
to the XOR condition (sample sizes differ due to random assignment). Stimuli
were squares varying in color (grayscale 9.8\%--90.2\%) and side length
(3.0--5.8cm), see Figure \ref{fig:e1-conditions}. The assignment of perceptual
features (color, size) to axes of the conceptual space (X, Y)  and the
direction of variation along each axis (e.g., $dark \rightarrow light$ or $light
\rightarrow dark$) were counterbalanced across participants.

@@

A secondary goal of this experiment was to examine whether we replicate the
classic result that generated categories often possess the same distributional
properties as previously-known categories. Given the increased importance of
replication within psychology \citep{zwaan2018making}, it is important as it
serves as a conceptual replication of \citet{jern2013probabilistic}. For each
generated category, we computed the category range along each axis (X, Y), as
well as the correlation between features. These data, shown in Figure
\ref{fig:e1-statsboxes}, reveal broad individual differences: Within each
condition, some participants generated categories spanning the entire X- and Y-
axis whereas other participants categories that spanned very little along each.
Likewise, in each condition participants generated categories possessing
strongly positive, neutral, and strongly negative correlations between the
dimensions. Comparing the distributional statistics between conditions yields a
broad yet, as we will see, misleading replication of the classic effect.

@@

However, our ability to detect differences in Y-axis range using a standard
$t$-test between the conditions is, in this case, diminished due to the
non-normality of the data (Shapiro-Wilk normality test $W=0.77, p < .001$ for
the Middle condition and $W=0.85, p < .001$ for the Bottom condition). Figure \ref{fig:e2-yranges} depicts the Y-axis range and Y-axis position of exemplars generated by each participant. The categories are sorted by overall range, and then colored by training condition. These data reveal that there were nearly as many
participants who generated categories spanning the entire Y-axis as those who
generated categories spanning almost none of the Y-axis. Indeed, the median produced Y-axis range is much smaller in the Bottom than Middle condition, whereas they are essentially identical in X-axis range. The non-normality of the Y-axis range distributions thus requires that we use a different analytic approach to
addressing the experiment's main question.

\begin{figure}
    \begin{center} \inputpgf{figs/}{e2-yranges.pgf}
    \caption{Y-axis range and position of the participant-generated categories
from Experiment 2. Each line corresponds to a participant's category, with
notches corresponding to the Y-axis position of exemplars within the category
(notches may overlap). Participants are sorted by overall range, and then by
condition. }
    \label{fig:e2-yranges}
    \end{center}
\end{figure}

@@

To obtain a global measure of the quality of each model's account, we fit the
parameters of each model to our entire dataset (Experiments 1 and 2 combined),
using a hill-climbing algorithm which maximized the log-likelihood of the
model's predictions of the observed responses (1220 responses from 305 total
participants). We fit three parameters in the PACKER model ($c$, $\theta_t$,
and $\theta_c$; see Section \ref{section:PACKER-definition}), as well as four in
the representativeness model and the hierarchical Bayesian model ($\kappa$,
$\lambda$, $\nu$, and $\theta$; see Section
\ref{section:representativeness-definition} and Appendix
\ref{ap:hsampling-definition} respectively). We fit only two parameters for
the copy-and-tweak model ($c$, and $\theta_t$), as $\theta_c$ is held constant
($\theta_c = 0$). Attention ($w$, see Equation \ref{eq:similarity}) in PACKER
and copy-and-tweak was set uniformly. Parameters were not allowed to vary
between participants or conditions -- the goal was to obtain the best-fitting
values to our entire dataset.

@@

While PACKER's advantage over copy-and-tweak may tentatively be attributed to
the model's sensitivity to category contrast (this will be explored in detail
below), the advantage shown by copy-and-tweak over the hierarchical Bayesian
model may be attributed to its exemplar-based representation of category B, as
opposed to forcing a prototype-based representation as assumed by the
hierarchical Bayesian model. As observed in Figures \ref{fig:e1-samples} and
\ref{fig:e2-samples}, the generated categories we observed were often widely
distributed, with no items near the category prototype. This aspect of the data
is inconsistent with the multivariate normal distributions (similar to
prototypes) used to represent categories in the \cite{jern2013probabilistic}
model, but can be easily accounted for using an exemplar-based approach.
Interestingly, representativeness using a prototype approach fits better than an
exemplar-based approach.

@@

In this section, we focus on a different category type -- the Corner category --
that contrasts the two models of contrast. Here, each Alpha exemplar is located
at the corners of the feature space (see Figure \ref{fig:e3-squares}a). We first
identify the predictions of each model in this condition. Second, we test these
contrasting predictions in a behavioral experiment. We then conduct a
model-based analysis of the two contrast models.

@@

Thus, in this experiment, we also examine how making one feature toroidal (e.g.,
line orientation) affects the prediction of each model. The representativeness
model prefers to generate categories on the feature that still has a boundary.
PACKER, in contrast, should continue to generate categories that are most
distant from the learn category regardless of the presence or absence of feature
boundaries -- in this condition it would be the center of the bounded feature.
Experiment 3 manipulates whether one feature is bounded or toroidal (always
displayed as the Y-axis). One condition uses the same stimuli as before, squares
varying in lightness and size, both of which are bounded features. The other
condition use a new set of stimuli: circles that vary in size (a bounded
feature) and orientation (an unbounded feature; Figure \ref{fig:e3-circles}a).
Before describing the details of the experiment, we explain in detail each
model's predictions given the Corner Alphas for a feature space where one
feature is bounded and the other is toroidal.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-circles.pdf}
      \caption{Circle stimulus domain (a) and predictions from contrast models
        PACKER (b) and the representativeness model (c). The unbounded feature
        (orientation) is represented on the Y-axis, while the bounded feature
        (size) is represented on a X-axis. As with the square stimuli, feature
        and direction assignment were counterbalanced across participants (in
        the displayed example, circles are rotated clockwise up the Y-axis). }
      \label{fig:e3-circles}
    \end{center}
\end{figure}

@@

For the representativeness model, adding a toroidal feature changes the
probability distribution for the underlying Bayesian model. For the toroidal
feature, space wraps back to $0$ radians once it reaches $2\pi$ (we have
conducted a change of variables where $y \rightarrow y$ mod $2\pi$, where mod is
the modulus operator). As we cross $2\pi$ and reach $0$ again, the remaining
probability density also gets wrapped around, meaning that the density at $2\pi$
is added to $0$ (and $2\pi + \epsilon$ to $\epsilon$, and so on). This is called
the {\em wrapped} Gaussian distribution \citep{mardia1972}. Note that the more
familiar distribution for circular features, the Von Mises distribution, is an
approximation to the wrapped Gaussian distribution. See Appendix B for
implementation details of the representativeness model in a bounded and
toroidal feature space.

@@

@@

We recruited 89 participants from Amazon Mechanical Turk who were randomly
assigned to the square stimuli ($N=46$), or to the circle stimuli ($N=43$). The
square stimuli were exactly the same as in Experiments 1 and 2, but with Alpha
exemplars in the corners of the feature space. The circle stimuli were made up
of unfilled circles with a radial line extending from its center to its edge.
These stimuli varied along two features: their diameter within the range of
[3.0, 5.8] cm (inclusive), and their orientation in the range [$2.587$,
$2\pi + 2.587$] radians (inclusive), with $0$ radians indicating a completely
horizontal line and orientation increasing in an anti-clockwise direction. The
line widths of the circles and radial line were set to 0.05 cm. Similar to
Experiments 1 and 2, participants first observed a training phase where four
Alpha exemplars were presented across three blocks followed by the generation
phase, and the features values are divided into 9 discrete steps. The training
condition was the Corner category for all participants.

@@

The distributions of generated first-Betas (Figure \ref{fig:e3-behave}) indicate
further qualitative support for PACKER. In particular, the probability of
generating the first Beta exemplar for circles along the bounded dimension is
significantly negatively correlated with distance from the center
($r(3) = -.95, p = .01$). Although the correlations for squares
($r(3)= -.50, p = .40$ and $r(3)= -.31, p = .61$ for X and Y dimensions
respectively) were not significant, they were in the directions predicted by
PACKER. Mean probabilities of generating first-Betas are presented in Figure
\ref{fig:e3-distcent}. This analysis was excluded for the toroidal dimension of
the circle stimuli because the boundaries (and consequently a center) are not
defined for this feature.

@@

\begin{figure}
    \begin{center}
      \includegraphics[width=.6\textwidth]{figs/e3-distcent.pdf}
      \caption{Empirical mean probabilities of generating first Beta exemplars
        at each point of distance from the center of the feature space.
        Distances are normalized such that 0 represented the center and 1
        represents the edge. Plot for the toroidal feature of circle stimuli is
        not presented because a center is not defined for this feature.}
    \label{fig:e3-distcent}
    \end{center}
\end{figure}
